---
title: "Differential gene expression analysis for Ctrl and Notch KO data"
author: "Nick Hirschmüller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
set.seed(123)
```


## Introduction
Run differential gene expression analysis for different celltypes using `muscat`. 
To run the differential expression analysis, we will use the "predicted" labels from MELD. I.e. only compare "truely" perturbed cells with "truely" unperturbed cells.


```{r}
library(Seurat)
library(tidyverse)
library(here)
library(muscat)
library(SingleCellExperiment)
library(DGEvistools)
library(patchwork)
library(PCAtools)
library(DESeq2)

source(here("plot_theme.R"))
source(here("helper_functions.R"))
```


### Coarse Annotation
Do the differential expression for a very coarse annotation (just 3 celltypes).

```{r}
seurat <- readRDS(here("output", "Ctrl_NotchKO_progenitors_integrated_meld.rds"))
seurat$VFC_2 <- ifelse(seurat$VFC_2 == 0, "ctrl_like", "notch_like")

# remove cells that come from pertubrbation but have label ctrl_like
mask1 <- seurat$perturbation == "notch" & seurat$VFC_2 == "ctrl_like"

# remove cells that come from ctrl but have label notch_like
mask2 <- seurat$perturbation == "ctrl" & seurat$VFC_2 == "notch_like"

# this contains no "missclassified" cells.
sce <- as.SingleCellExperiment(seurat[, !(mask1 | mask2)])

# muscat expects certain columns:
sce$sample_id <- factor(sce$orig.ident, levels = c("TX22", "TX23", "TX22_N", "TX23_N"))
sce$cluster_id <- "Progenitor"
sce$group_id <- factor(sce$perturbation, levels = c("ctrl", "notch"))

# prepare the sce object
sce <- prepSCE(sce, drop = T)

# number of cells per cluster-sample
t(table(sce$cluster_id, sce$sample_id))


# this does a different assay for each celltype
pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id")
)


# construct design & contrast matrix
ei <- metadata(sce)$experiment_info
ei$batch <- c("TX22", "TX23", "TX22", "TX23")
mm <- model.matrix(~ batch + group_id, data = ei)
dimnames(mm) <- list(ei$sample_id, c("ctrl", "TX23", "notch"))
contr <- limma::makeContrasts("notch", levels = mm) # compare notch VS ctrl

res_edgeR <- pbDS(pb, design = mm, method = "edgeR", contrast = contr)
```

# save edgeR results
```{r}
for (celltype in names(res_edgeR$table$notch)) {
    dir.create(here("output", "differential_expression", "pseudobulk_coarse_meld", celltype), showWarnings = F, recursive = T)
    res_edgeR$table$notch[[celltype]] %>%
        dplyr::select(gene, logCPM, logFC, p_adj = p_adj.loc, celltype = cluster_id) %>%
        arrange(p_adj) %>%
        data.table::fwrite(.,
            file = here("output", "differential_expression", "pseudobulk_coarse_meld", celltype, "edgeR_diff_expr.tsv"), sep = "\t",
            row.names = F
        )
}
```

### Granular Annotation
Do the differential expression for a more granular annotation.

```{r}
seurat <- readRDS(here("output", "Ctrl_NotchKO_progenitors_integrated_meld.rds"))
seurat$VFC_2 <- ifelse(seurat$VFC_2 == 0, "ctrl_like", "notch_like")

# remove cells that come from pertubrbation but have label ctrl_like
mask1 <- seurat$perturbation == "notch" & seurat$VFC_2 == "ctrl_like"

# remove cells that come from ctrl but have label notch_like
mask2 <- seurat$perturbation == "ctrl" & seurat$VFC_2 == "notch_like"

# this contains no "missclassified" cells.
sce <- as.SingleCellExperiment(seurat[, !(mask1 | mask2)])

# muscat expects certain columns:
sce$sample_id <- factor(sce$orig.ident, levels = c("TX22", "TX23", "TX22_N", "TX23_N"))
sce$cluster_id <- sce$high_res_annotation
sce$group_id <- factor(sce$perturbation, levels = c("ctrl", "notch"))

# prepare the sce object
sce <- prepSCE(sce, drop = T)


# number of cells per cluster-sample
t(table(sce$cluster_id, sce$sample_id))


# this does a different assay for each celltype
pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id")
)


# construct design & contrast matrix
ei <- metadata(sce)$experiment_info
ei$batch <- c("TX22", "TX23", "TX22", "TX23")
mm <- model.matrix(~ batch + group_id, data = ei)
dimnames(mm) <- list(ei$sample_id, c("ctrl", "TX23", "notch"))
contr <- limma::makeContrasts("notch", levels = mm) # compare notch VS ctrl

res_edgeR <- pbDS(pb, design = mm, method = "edgeR", contrast = contr)
```

# save edgeR results
```{r}
for (celltype in names(res_edgeR$table$notch)) {
    dir.create(here("output", "differential_expression", "pseudobulk_granular_meld", celltype), showWarnings = F, recursive = T)
    res_edgeR$table$notch[[celltype]] %>%
        dplyr::select(gene, logCPM, logFC, p_adj = p_adj.loc, celltype = cluster_id) %>%
        arrange(p_adj) %>%
        data.table::fwrite(.,
            file = here("output", "differential_expression", "pseudobulk_granular_meld", celltype, "edgeR_diff_expr.tsv"), sep = "\t",
            row.names = F
        )
}
```


```{r}
sessionInfo()
```





