---
title: "Trajectory inference using conditments"
author: "Nick Hirschmüller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
set.seed(123)
```


## Introduction
Using the trajectory inference methods described [here](https://www.biorxiv.org/content/10.1101/2021.03.09.433671v1.full).
We willl do this for the KO dataset


```{r}
library(Seurat)
library(tidyverse)
library(patchwork)
library(here)
library(slingshot)
library(tradeSeq)
library(scran)
library(condiments)

# source plotting theme
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```

```{r}
# load seurat object
seurat <- readRDS(here("output", "Ctrl_NotchKO_integrated_scent.rds"))

# specify the celltypes we want to include in our analysis.
seurat <- seurat[, !seurat$celltype_manual %in% c("MT", "unk", "unk2")]

# regroup some celltypes
seurat@meta.data$tmp_celltype <- case_when(
  seurat$high_res_annotation == "EEP" ~ "EEP", # include the EE progenitor population
  seurat$celltype_manual %in% c("dEC", "daEC") ~ "dEC",
  seurat$celltype_manual %in% c("LFC", "Copper") ~ "mEC",
  T ~ seurat$celltype_manual
)

# because we have removed some celltypes rerun integration
seurat_split <- SplitObject(DietSeurat(seurat, assays = "RNA"), split.by = "orig.ident")
seurat_split <- lapply(seurat_split,function(x){
  x <- NormalizeData(x, normalization.method = "LogNormalize", scale.factor = 10000)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 3000)
  x <- ScaleData(x)
  return(x)
})

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = seurat_split)

# find anchors between experiments
anchors <- FindIntegrationAnchors(object.list = seurat_split,
                                  anchor.features = features)

seurat <- IntegrateData(anchorset = anchors,
                                          dims = 1:15)

DefaultAssay(seurat) <- "integrated"
seurat <- ScaleData(seurat, verbose = F) %>%
  RunPCA(., npcs = 50, verbose = F) %>%
  RunUMAP(., reduction = "pca", dims = 1:20, verbose = F)
DefaultAssay(seurat) <- "RNA"
saveRDS(seurat, here("output", "seuratKO_slingshot_object_integrated.rds"))
```

```{r}
# prepare the slingshot input
# just to make visually more pleasing (diff from left to right)
seurat[["pca_mod"]] <- CreateDimReducObject(
    embeddings = Embeddings(seurat, reduction = "pca") * -1,
    key = "PCAmod_",
    assay = "integrated"
)

# create sce object
DefaultAssay(seurat) <- "RNA"
seurat_sce <- as.SingleCellExperiment(seurat)
reducedDim(seurat_sce, "PCA") <- Embeddings(seurat, "pca_mod")[, 1:30]

slingshot_res <- slingshot(seurat_sce,
    reducedDim = "PCA",
    clusterLabels = colData(seurat_sce)$tmp_celltype,
    start.clus = c("ISC"),
    end.clus = c("aEC", "mEC", "pEC"),
    approx_points = 150
)
reducedDim(slingshot_res, "UMAP") <- Embeddings(seurat, "umap")
saveRDS(slingshot_res, here("output", "slingshot_object.rds"))
```

### Refine fit
Evaluate if the fit was successful or if we should fit separate trajectories per condition.
```{r}
# Authors of condiments/slingshot prefer fitting a common trajectory if the differences
# between the conditions is small enough!
# here we check if this is the case for our data
slingshot_res <- readRDS(here("output", "slingshot_object.rds"))

scores <- condiments::imbalance_score(
    Object = Embeddings(seurat, "pca_mod")[, 1:2] %>% as.matrix(),
    conditions = seurat$perturbation,
    k = 40, smooth = 30
)

seurat$scores <- scores$scaled_scores
imbalence_df <- data.frame(Embeddings(seurat, "pca_mod")[, 1:2])
imbalence_df$scores <- scores$scaled_scores

# we can see, that there are areas of great imbalence. Which is expected given we see different celltype abundance
ggplot(imbalence_df, aes(x = PCAmod_1, y = PCAmod_2, col = scores)) +
    geom_point(size = .7) +
    scale_color_viridis_c(option = "C") +
    labs(col = "Scores")


# essentially, the labels get permutated and the trajectories are refit.
# pseudotime vector is generated and the distribution is compared with the "true" one
# if the results do not change significantly, we can fit the same trajectory for both conditions.

# This result suggests we should fit seperat trajetories for the conditions
topologyTest_result <- topologyTest(SlingshotDataSet(slingshot_res),
    slingshot_res$perturbation,
    rep = 100,
    methods = "KS_mean",
    threshs = .01,
    parallel = TRUE,
    BPPARAM = BiocParallel::MulticoreParam(workers = 50)
)


# refit slingshot
slingshot_individual <- slingshot_conditions(slingshot_res,
    slingshot_res$perturbation,
    approx_points = 150
)
saveRDS(slingshot_individual, here("output", "slingshot_individual_fit.rds"))
```


## Evaluate the fit
Despite fitting separate trajectories, we still observe that a common skeleton for the minimal spanning tree exists, so we can merge them (see Fig1e of the condiments paper [https://www.biorxiv.org/content/10.1101/2021.03.09.433671v1.full])

```{r}
slingshot_individual <- readRDS(here("output", "slingshot_individual_fit.rds"))
slingshot_individual$condition_id <- names(slingshot_individual)
slingshot_individual$mapping <- matrix(rep(1:4, each = 2), nrow = 4, ncol = 2, byrow = TRUE)
slingshot_merged <- do.call(merge_sds, slingshot_individual)


#################
# VISUALIZATION #
#################

# show that the trajectories are very similar

plot_df <- Embeddings(seurat, "pca_mod")[, c(1, 2)] %>%
    data.frame() %>%
    rownames_to_column("cells")

plot_df$condition <- seurat@meta.data$perturbation
plot_df$celltype <- seurat@meta.data$tmp_celltype

plot_df <- plot_df %>%
    left_join(., slingPseudotime(slingshot_merged) %>%
        as.data.frame() %>%
        mutate(cells = rownames(.)),
    by = "cells"
    ) %>%
    pivot_longer(starts_with("Lineage"), names_to = "Curve", values_to = "pst") %>%
    sample_frac(1)


p <- ggplot(plot_df, aes(x = PCAmod_1, y = PCAmod_2, col = condition)) +
    geom_point(size = .1, alpha = 0.3) +
    small_axis("UMAP ") +
    xlab("PC 1") +
    ylab("PC 2")

# follow very similar trajectories!
for (condition in c("ctrl", "notch")) {
    sds_cond <- slingshot_individual[[condition]]
    for (i in 1:4) {
        p <- p +
            geom_path(
                data = slingCurves(sds_cond)[[i]]$s[slingCurves(sds_cond)[[i]]$ord, ] %>%
                    as.data.frame() %>%
                    mutate(condition = condition),
                linewidth = 1
            )
    }
}
p

# calculate the cluster centroids per condition
cls <- plot_df %>%
    group_by(celltype, condition) %>%
    dplyr::summarise(PCAmod_1 = mean(PCAmod_1), PCAmod_2 = mean(PCAmod_2))

p <- ggplot(plot_df, aes(x = PCAmod_1, y = PCAmod_2, col = condition)) +
    geom_point(size = .1, alpha = 0.3) +
    small_axis("UMAP ") +
    xlab("PC 1") +
    ylab("PC 2")

p <- p +
    geom_point(data = cls, size = 2.1)

edges <- lapply(slingLineages(slingshot_merged), function(lin) {
    from <- lin[1:(length(lin) - 1)]
    to <- lin[2:length(lin)]
    return(data.frame("from" = from, "to" = to))
}) %>% bind_rows()

for (condition_ in c("ctrl", "notch")) {
    cl_condition <- cls %>% filter(condition == condition_)
    edges_condition <- left_join(edges, cl_condition, by = c("from" = "celltype")) %>%
        left_join(cl_condition %>% dplyr::rename(
            "PCAmod_1_end" = "PCAmod_1",
            "PCAmod_2_end" = "PCAmod_2"
        ) %>%
            select(-condition), by = c("to" = "celltype"))
    p <- p +
        geom_segment(
            data = edges_condition, aes(xend = PCAmod_1_end, yend = PCAmod_2_end),
            size = 1
        )
}
p
```

## Run tradeSeq
```{r}
# find all genes, that have at least 10 total counts in 5 different cells.
geneFilter <- apply(as.matrix(counts(slingshot_res)), 1, function(x) {
    sum(x >= 5) >= 5
})

slingshot_res <- slingshot_res[geneFilter, ]
slingshot_res@int_metadata$slingshot <- slingshot_merged

# this takes very long, even when using multiple cores
# library(BiocParallel)
# BPPARAM <- BiocParallel::bpparam()
# BPPARAM$workers <- 60
gam_fit_6 <- fitGAM(
    counts = slingshot_res,
    conditions = factor(colData(slingshot_res)$perturbation),
    parallel = F,
    # BPPARAM = BPPARAM,
    verbose = T,
    nknots = 6
)
saveRDS(gam_fit_6, here("output", "tradeSeq_fit_condiments.rds"))

gam_fit_6 <- readRDS(here("output", "tradeSeq_fit_condiments.rds")) 
```


```{r}
assoc_results <- tradeSeq::associationTest(gam_fit_6, lineages = TRUE, l2fc = log2(2), global = FALSE)

# divide the df into something more digestable
assoc_results_split <- lapply(split(1:24, rep(c(1, 2, 3, 4), each = 6)), function(x) {
    tmp <- assoc_results[, x] %>%
        rownames_to_column("gene") %>%
        dplyr::select(-contains("waldStat"), -contains("df_"))

    tmp$padj_ctrl <- p.adjust(tmp[, 2], "fdr")
    tmp$padj_notch <- p.adjust(tmp[, 3], "fdr")
    return(tmp[, c(1, 4, 5)])
})
names(assoc_results_split) <- c("Lineage1", "Lineage2", "Lineage3", "Lineage4")
saveRDS(assoc_results_split, here("output", "tradeSeq_assoc_res.rds"))


condRes <- conditionTest(gam_fit_6, l2fc = log2(2), global = FALSE, lineages = TRUE)
# divide the df into something more digestable
condRes_split <- lapply(split(1:12, rep(c(1, 2, 3, 4), each = 3)), function(x) {
    tmp <- condRes[, x] %>%
        rownames_to_column("gene") %>%
        dplyr::select(-contains("waldStat"), -contains("df_"))

    lin <- str_extract(colnames(tmp)[2], "lineage\\d")

    tmp[[paste0("padj_", lin)]] <- p.adjust(tmp[, 2], "fdr")
    return(tmp[, c(1, 3)])
})
names(condRes_split) <- c("Lineage1", "Lineage2", "Lineage3", "Lineage4")
saveRDS(condRes_split, here("output", "tradeSeq_condition_test_res.rds"))
```


```{r}
sessionInfo()
```










