---
title: "Differential gene expression analysis for CphUp vs Ctrl"
author: "Nick Hirschmüller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
set.seed(123)
```


## Introduction
Run differential gene expression analysis for all subtypes using `muscat`.


```{r}
library(Seurat)
library(tidyverse)
library(here)
library(muscat)
library(SingleCellExperiment)
library(scater)
library(ggVennDiagram)
library(DGEvistools)
library(patchwork)
library(clusterProfiler)
library(PCAtools)
library(DESeq2)

source(here("plot_theme.R"))
source(here("helper_functions.R"))
```


We include the 3 primes because why not, just increases runtime. And we cant use the globally adjusted p values than, but we take local ones anyway. 
Upon reflection, we will remove them. We are not really interested in differences there, and 
information across samples is used during differential gene expression and also during clustering.

```{r}
seurat <- readRDS(here("output", "RNAi_experiments_integrated.rds"))

seurat <- seurat[, seurat$perturbation %in% c("ctrl", "CphUp")]


# we want comparision for ISC and EEs and ISC+EEPs (will be essentially the same as ISC because almost no EEPs in ctrl)
seurat <- seurat[, !seurat$celltype_manual %in% c("unk", "unk2", "MT", "dEC", "daEC", "EB")]
seurat <- seurat[, !seurat$high_res_annotation == "EEP"]

seurat@meta.data$tmp_celltype <- case_when(
    seurat$celltype_manual %in% c("mEC", "aEC", "pEC", "LFC", "Copper") ~ "EC",
    T ~ seurat$celltype_manual
)

sce <- as.SingleCellExperiment(seurat)

# muscat expects certain columns:
sce$sample_id <- factor(sce$orig.ident.unique, levels = sce$orig.ident.unique %>% unique())
sce$cluster_id <- sce$tmp_celltype
sce$group_id <- factor(sce$perturbation, levels = sce$perturbation %>% unique())
table(sce$sample_id, sce$cluster_id)

# prepare the sce object
sce <- prepSCE(sce, drop = T)
```


## PCA based QC
```{r, eval=F}
###########################
# PCA USING ALL CELLTYPES #
###########################
# this combines all celltypes in one assay. Better if we want to visulaize stuff.
whole_pseudobulk <- glmGamPoi::pseudobulk(sce,
    group_by = vars(sample_id, cluster_id, group_id)
)

# plot a PCA
vst_data <- DESeq2::vst(counts(whole_pseudobulk))


vst_corrected <- limma::removeBatchEffect(vst_data,
    batch = substr(colnames(vst_data), 1, 4)
)


pattern <- "([^.]+)\\.([^.]+)\\.([^.]*)"
sample_mdata <- data.frame(do.call(rbind, lapply(str_match_all(colnames(vst_corrected), pattern), function(x) x[, 2:4])))
colnames(sample_mdata) <- c("sample_id", "celltype", "perturbation")
rownames(sample_mdata) <- colnames(vst_corrected)
sample_mdata$batch <- str_extract(sample_mdata$sample_id, "TX\\d+")

stopifnot(all(rownames(sample_mdata) == colnames(vst_corrected)))

# run PCA using 60% must variable genes
pca_data <- pca(vst_corrected,
    metadata = sample_mdata,
    removeVar = 0.4
)

# extract top 6 PCS
pca_plot_df <- pca_data$rotated[, c("PC1", "PC2", "PC3")] %>%
    rownames_to_column("sample_id") %>%
    left_join(., sample_mdata %>% rownames_to_column("tmp"), by = c("sample_id" = "tmp"))


# looks really nice actually
p <- ggplot(pca_plot_df, aes(PC1, PC2, color = perturbation, shape = batch, size = celltype)) +
    geom_point() +
    xlab(paste0("PC1: ", round(pca_data$variance[1], 2), "% variance")) +
    ylab(paste0("PC2: ", round(pca_data$variance[2], 2), "% variance")) +
    theme_Publication() +
    scale_size_manual(values = c(2, 5, 10)) +
    chameleon::scale_color_chameleon()



celltype_pca <- lapply(c("EE", "EC", "ISC"), function(x) {
    mask <- grepl(x, colnames(counts(whole_pseudobulk)))
    vst_data <- DESeq2::vst(counts(whole_pseudobulk)[, mask])
    vst_corrected <- limma::removeBatchEffect(vst_data,
        batch = substr(colnames(vst_data), 1, 4)
    )

    pattern <- "([^.]+)\\.([^.]+)\\.([^.]*)"

    sample_mdata <- data.frame(do.call(rbind, lapply(str_match_all(colnames(vst_corrected), pattern), function(x) x[, 2:4])))
    colnames(sample_mdata) <- c("sample_id", "celltype", "perturbation")
    rownames(sample_mdata) <- colnames(vst_corrected)
    sample_mdata$batch <- str_extract(sample_mdata$sample_id, "TX\\d+")

    stopifnot(all(rownames(sample_mdata) == colnames(vst_corrected)))

    # run PCA using 60% must variable genes
    pca_data <- pca(vst_corrected,
        metadata = sample_mdata,
        removeVar = 0.4
    )

    # extract top 6 PCS
    pca_plot_df <- pca_data$rotated[, c("PC1", "PC2", "PC3")] %>%
        rownames_to_column("sample_id") %>%
        left_join(., sample_mdata %>% rownames_to_column("tmp"), by = c("sample_id" = "tmp"))

    ggplot(pca_plot_df, aes(PC1, PC2, color = perturbation, shape = batch)) +
        geom_point(size = 4) +
        xlab(paste0("PC1: ", round(pca_data$variance[1], 2), "% variance")) +
        ylab(paste0("PC2: ", round(pca_data$variance[2], 2), "% variance")) +
        theme_Publication() +
        chameleon::scale_color_chameleon() +
        ggtitle(x) +
        guides(color = guide_legend(nrow = 2, byrow = TRUE), shape = guide_legend(nrow = 2, byrow = TRUE))
})

p <- wrap_plots(celltype_pca, nrow = 1)
p
```





### Run differential expression 
```{r}
###############################
# RUN DIFFERENTIAL EXPRESSION #
###############################
colData(sce)$sample_id <- sce$sample_id %>% droplevels()
colData(sce)$group_id <- sce$group_id %>% droplevels()
colData(sce)$cluster_id <- sce$cluster_id %>% droplevels()
table(sce$sample_id, sce$cluster_id)


# this does a different assay for each celltype
pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id")
)

# one sheet per subpopulation
assayNames(pb)

# construct design & contrast matrix
ei <- metadata(sce)$experiment_info 

ei$batch <- substr(ei$sample_id, 1, 4)
mm <- model.matrix(~ batch + group_id, data = ei)
dimnames(mm) <- list(ei$sample_id, c("Ctrl", "TX49", "CphUp"))

# this corresponds to the difference of CphUp vs ctrl is our baseline
# also explained here: https://support.bioconductor.org/p/110759/
contr <- limma::makeContrasts("CphUp", levels = mm) # compare notch VS ctrl

res <- pbDS(pb,
    design = mm,
    method = "edgeR",
    contrast = contr,
    min_cells = 0,
    filter="genes"
)

```

# save edgeR results
```{r}
for (celltype in names(res$table$CphUp)) {
    dir.create(here("output", "differential_expression", "pseudobulk_CphUp", celltype), showWarnings = F, recursive = T)
    res$table$CphUp[[celltype]] %>%
        dplyr::select(gene, logCPM, logFC, p=p_val, p_adj = p_adj.loc, celltype = cluster_id) %>%
        arrange(p_adj) %>%
        data.table::fwrite(.,
            file = here("output", "differential_expression", "pseudobulk_CphUp", celltype, "edgeR_diff_expr.tsv"), sep = "\t",
            row.names = F
        )
}
```

```{r}
sessionInfo()
```

