---
title: "Differential gene expression analysis for RNAi data using FindMarkers"
author: "Nick Hirschmüller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
set.seed(123)
```


## Introduction
For NotchRNAi and NotchRNAi+CphRNAi we only have one replicate so cant do pseudobulk... instead use the conventional FindMarkers to get an idea. Previous reports have shown that this leads to many false positives, so interpretation of p values should be done with care!

We will do the differential expression for mulitple populations:
ISC
EB
EEP
ISC+EEP
ISC+EEP+EB

And also for multiple contrasts: 
NotchRNAi vs ctrl
NotchRNAi+CphRNAi vs ctrl
NotchRNAi vs NotchRNAi+CphRNAi

```{r}
library(Seurat)
library(tidyverse)
library(here)
library(patchwork)

source(here("plot_theme.R"))
source(here("helper_functions.R"))

seurat <- readRDS(here("output", "Ctrl_Notch_NotchCphRNAi_integrated_scent.rds"))
DefaultAssay(seurat) <- "RNA"
```


### Run Find Markers

```{r}
celltypes <- list(
    c("ISC","EB"),
    c("ISC"),
    c("EB"),
    c("EEP"),
    c("ISC", "EEP"),
    c("ISC", "EEP", "EB")
)

comparisions <- list(
    c("NotchRNAi", "ctrl"),
    c("NotchCphRNAi", "ctrl"),
    c("NotchRNAi", "NotchCphRNAi")
)

for (celltype in celltypes) {
    celltypes_string <- paste(celltype, collapse = "_")
    print(celltypes_string)
    tmp_seurat <- seurat[, seurat$high_res_annotation %in% celltype]
    DefaultAssay(tmp_seurat) <- "RNA"
    Idents(tmp_seurat) <- tmp_seurat$perturbation
    for (comp in comparisions) {
        print(str_interp("Running comparision ${comp[1]} vs ${comp[2]}"))
        if(all(celltype == "EEP")) {
            if ("NotchCphRNAi" %in% comp){
                print("skipping")
                next()
            }
        }
        result <- FindMarkers(
            tmp_seurat,
            ident.1 = comp[1],
            ident.2 = comp[2],
            test.use = "MAST", # use MAST to account for cellular detection rate
            logfc.threshold = 0.01, # this increases runtime, and marginally increases p values.
            min.pct = 0.01
        ) %>% rownames_to_column("gene") %>% 
            mutate(celltype=celltypes_string) %>% 
            mutate(comparision = str_interp("${comp[1]}_vs_${comp[2]}"))
        
        dir.create(here("output","differential_expression","FindMarkers_MAST",celltypes_string), showWarnings = FALSE)
        data.table::fwrite(result, 
                           file = here("output","differential_expression","FindMarkers_MAST",celltypes_string, 
                                       str_interp("${comp[1]}_vs_${comp[2]}_${celltypes_string}_MAST.tsv")),
                           sep="\t",
                           row.names = F)
    }
}


```



```{r}
sessionInfo()
```

