---
title: "Differential gene expression analysis for Ctrl and Notch KO data"
author: "Nick Hirschmüller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
set.seed(123)
```


## Introduction
Run differential gene expression analysis for different celltypes using `muscat`. 
Here we will use the conventional pseudobulk approach. I.e. we ignore the labels that have been predicted by MELD.



```{r}
library(Seurat)
library(tidyverse)
library(here)
library(muscat)
library(SingleCellExperiment)
library(DGEvistools)
library(patchwork)
library(PCAtools)
library(DESeq2)

source(here("plot_theme.R"))
source(here("helper_functions.R"))
```


### Coarse Annotation
Do the differential expression for a very coarse annotation (just 3 celltypes).

```{r}
seurat <- readRDS(here("output", "Ctrl_NotchKO_per_celltype_res_meld.rds"))

# remove celltypes that we are not interested in, or which do not occure in all samples
seurat <- seurat[, !seurat$celltype_manual %in% c("unk", "unk2", "MT", "dEC", "daEC")]

# aggregate different related  celltypes:

# Progenitors -> EEP, EB, ISC
# EE -> EE
# EC -> aEC, pEC, mEC, Copper, LFC
seurat@meta.data$tmp_celltype <- case_when(
    seurat$celltype_manual %in% c("ISC", "EB") ~ "Progenitor",
    seurat$high_res_annotation == "EEP" ~ "Progenitor",
    grepl("-EE", seurat$high_res_annotation) ~ "EE",
    seurat$celltype_manual %in% c("mEC", "aEC", "pEC", "LFC", "Copper") ~ "EC",
    T ~ seurat$celltype_manual
)
sce <- as.SingleCellExperiment(seurat)

# muscat expects certain columns:
sce$sample_id <- factor(sce$orig.ident, levels = c("TX22", "TX23", "TX22_N", "TX23_N"))
sce$cluster_id <- sce$tmp_celltype
sce$group_id <- factor(sce$perturbation, levels = c("ctrl", "notch"))

# prepare the sce object
sce <- prepSCE(sce, drop = T)

# number of cells per cluster-sample
t(table(sce$cluster_id, sce$sample_id))


# this does a different assay for each celltype
pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id")
)


# construct design & contrast matrix
ei <- metadata(sce)$experiment_info
ei$batch <- c("TX22", "TX23", "TX22", "TX23")
mm <- model.matrix(~ batch + group_id, data = ei)
dimnames(mm) <- list(ei$sample_id, c("ctrl", "TX23", "notch"))
contr <- limma::makeContrasts("notch", levels = mm) # compare notch VS ctrl

res_edgeR <- pbDS(pb, design = mm, method = "edgeR", contrast = contr)
```

# save edgeR results
```{r}
for (celltype in names(res_edgeR$table$notch)) {
    dir.create(here("output", "differential_expression", "pseudobulk_coarse", celltype), showWarnings = F, recursive = T)
    res_edgeR$table$notch[[celltype]] %>%
        dplyr::select(gene, logCPM, logFC, p_adj = p_adj.loc, celltype = cluster_id) %>%
        arrange(p_adj) %>%
        data.table::fwrite(.,
            file = here("output", "differential_expression", "pseudobulk_coarse", celltype, "edgeR_diff_expr.tsv"), sep = "\t",
            row.names = F
        )
}
```

### Granular Annotation
Do the differential expression for a more granular annotation.

```{r}
seurat <- readRDS(here("output", "Ctrl_NotchKO_per_celltype_res_meld.rds"))

# remove celltypes that we are not interested in, or which do not occure in all samples
seurat <- seurat[, !seurat$celltype_manual %in% c("unk", "unk2", "MT", "dEC", "daEC")]

# aggregate different related  celltypes:
seurat@meta.data$tmp_celltype <- case_when(
    seurat$high_res_annotation == "EEP" ~ "EEP",
    grepl("-EE", seurat$high_res_annotation) ~ "EE",
    seurat$celltype_manual %in% c("mEC", "aEC", "pEC", "LFC", "Copper") ~ "EC",
    T ~ seurat$celltype_manual
)
sce <- as.SingleCellExperiment(seurat)

# muscat expects certain columns:
sce$sample_id <- factor(sce$orig.ident, levels = c("TX22", "TX23", "TX22_N", "TX23_N"))
sce$cluster_id <- sce$tmp_celltype
sce$group_id <- factor(sce$perturbation, levels = c("ctrl", "notch"))

# prepare the sce object
sce <- prepSCE(sce, drop = T)

# number of cells per cluster-sample
t(table(sce$cluster_id, sce$sample_id))


# this does a different assay for each celltype
pb <- aggregateData(sce,
    assay = "counts", fun = "sum",
    by = c("cluster_id", "sample_id")
)


# construct design & contrast matrix
ei <- metadata(sce)$experiment_info
ei$batch <- c("TX22", "TX23", "TX22", "TX23")
mm <- model.matrix(~ batch + group_id, data = ei)
dimnames(mm) <- list(ei$sample_id, c("ctrl", "TX23", "notch"))
contr <- limma::makeContrasts("notch", levels = mm) # compare notch VS ctrl

res_edgeR <- pbDS(pb, design = mm, method = "edgeR", contrast = contr)
```

# save edgeR results
```{r}
for (celltype in names(res_edgeR$table$notch)) {
    dir.create(here("output", "differential_expression", "pseudobulk_granular", celltype), showWarnings = F, recursive = T)
    res_edgeR$table$notch[[celltype]] %>%
        dplyr::select(gene, logCPM, logFC, p_adj = p_adj.loc, celltype = cluster_id) %>%
        arrange(p_adj) %>%
        data.table::fwrite(.,
            file = here("output", "differential_expression", "pseudobulk_granular", celltype, "edgeR_diff_expr.tsv"), sep = "\t",
            row.names = F
        )
}
```




```{r}
sessionInfo()
```





